\bf{Neural Ordinary Differential Equations}

### 2025-03-19 paper.tex
added derivation to report

\subsection{Sensitivity or Adjoint Equation Methods}
\label{sec:sens}

Here we give a quick overview of the sensitivity equations for ODEs. Consider following
non-linear state-space model:

\begin{align}
    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag \\
    y &= g(x, u; \theta)
\end{align}

@@

\begin{align}
    \min\limits_{\theta} V(x, u; \theta) = \norm{y - \hat{y}}^2 
\end{align}
To propagate the gradient we need to solve the sensitivity equations:
\begin{align}
    \dfrac{\partial V(x, u; \theta)}{\partial \theta_j} &= -2 \sum_i (\hat{x} - x) \dfrac{\partial x_i}{\partial \theta_j} \notag \\ 
    S_{ij} &= \dfrac{\partial x_i}{\partial \theta_j}
    \label{eq:sens}
\end{align}
where $S_{ij}$ is the sensitivity of model solution with parameter. We differentiate $S_{ij}$ to obtain:
\begin{align}
    \dfrac{d S_{ij}}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x_i}{\partial \theta_j}\right) =
    \dfrac{\partial}{\partial\theta_j} \left(\dfrac{d x_i}{dt}\right) = 
    \dfrac{\partial }{\partial \theta_j} f_i(x; \theta) 
\end{align}
Using chain rule on $f$, we get following matrix equation:
\begin{align}
    \dfrac{dS}{dt} = f_xS + f_{\theta} \qquad  S(0) = g_\theta
\end{align}
We could easily calculate the initial condition 
if we are using a state estimator by using the \cref{eq:sens}. Thus, we can construct
following augmented ODE system:
\begin{align}
    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
    \begin{bmatrix} f(x, u; \theta) \\ f_xS + f_{\theta} \end{bmatrix}
    \qquad
    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
    \begin{bmatrix} x_0 \\ g_\theta \end{bmatrix}
    \label{eq:augsens}
\end{align}
This augmented system can be solved using any ODE solver, allowing gradient 
propagation without the need to store intermediate states. This key advantage 
contributed to the popularity of Neural ODEs (NODEs), beyond their interpretation 
as a continuous counterpart to Euler-step-based architectures like ResNets. We close 
this section by a short example on solving the augmented ODE system. Consider, 
following kinetics:
\begin{align}
    \dot{c}_A &= -kc_A \notag \\
    c_A(0) &= c_{A0}
\end{align}
Say we want to calculate sensitivities wrt $k, c_{A0}$. With reference to \cref{eq:augsens},
for $S_1 = \dfrac{\partial c_A}{\partial k}$ and $S_2 = \dfrac{\partial c_A}{\partial c_{A0}}$,
we have following augmented system:
\begin{align}
    \begin{bmatrix} 
        \dot{c}_A \\
        \dot{S_1} \\ 
        \dot{S_2} 
    \end{bmatrix} = 
    \begin{bmatrix}
        -kc_A \\
        -kS_1 - c_A \\
        -kS_2 \end{bmatrix}
    \qquad
    \begin{bmatrix} 
        c_A(0) \\ 
        S_1(0) \\ 
        S_2(0) 
    \end{bmatrix} =
    \begin{bmatrix} 
        c_{A0} \\ 
        0 \\ 
        1
    \end{bmatrix}
\end{align}
If we were learning the parameter $k$ and initial condition $c_{A0}$, we could
show a gradient descent update as follows:
\begin{align}
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix} \leftarrow
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix}
    - \eta 
    \begin{bmatrix} 
        -2 (\hat{c}_A  - c_A)^T S_1 \\ 
        -2 (\hat{c}_A  - c_A)^T S_2
    \end{bmatrix}
\end{align}
Thus, we do no need to backprop but just solve the augmented ODE system once 
in each optimal iteration unlike ResNets or RNNs. However, such a method is not 
an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
numerical ODE solvers like stiffness, accuracy, etc. Sometimes we may end up 
with noisy gradients which may lead to
poor convergence.

>>>>

\subsection{Forward and backward Sensitivity/Adjoint methods}

\subsubsection{Forward Sensitivity}
\label{sec:sens}

Here we give a quick overview of both the forward/backward sensitivity equations for ODEs. Consider following
non-linear state-space model:

\begin{align}
    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag \\
    y &= g(x, u; \theta)
    \label{eq:ssode}
\end{align}

@@

\begin{align}
    \min\limits_{\theta} V(x) = \norm{x - \hat{x}}^2 
\end{align}
To propagate the gradient we need to solve the sensitivity equations:
\begin{align}
    \dfrac{\partial V(x)}{\partial \theta_j} &= -2 \sum_i (\hat{x} - x) \dfrac{\partial x_i}{\partial \theta_j} \notag \\ 
    S_{ij} &= \dfrac{\partial x_i}{\partial \theta_j}
    \label{eq:sens}
\end{align}
where $S_{ij}$ is the sensitivity of model solution with parameter. We differentiate $S_{ij}$ to obtain:
\begin{align}
    \dfrac{d S_{ij}}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x_i}{\partial \theta_j}\right) =
    \dfrac{\partial}{\partial\theta_j} \left(\dfrac{d x_i}{dt}\right) = 
    \dfrac{\partial }{\partial \theta_j} f_i(x; \theta) 
\end{align}
Using chain rule on $f$, we get following matrix equation:
\begin{align}
    \dfrac{dS}{dt} = f_xS + f_{\theta} \qquad  S(0) = g_\theta
\end{align}
We could easily calculate the initial condition 
if we are using a state estimator by using the \cref{eq:sens}. Thus, we can construct
following augmented ODE system:
\begin{align}
    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
    \begin{bmatrix} f(x, u; \theta) \\ f_xS + f_{\theta} \end{bmatrix}
    \qquad
    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
    \begin{bmatrix} x_0 \\ g_\theta \end{bmatrix}
    \label{eq:augsens}
\end{align}
This augmented system can be solved using any ODE solver, allowing gradient 
propagation. However, if $\theta \in \R^{p}$ and $x \in \R^{n}$, we need to 
solve $p \times n + 1$ equations i.e. the cost increases with $p$. Thus, forward
sensitivity is not an efficient method to propagate gradients. However, we can circumvent
the problem of storing intermediate states using backward sensitivity method discussed next. This key advantage 
of backward adjoint method contributed to the popularity of Neural ODEs (NODEs), beyond their interpretation 
as a continuous counterpart to Euler-step-based architectures like ResNets. We also show 
a short example on solving the augmented ODE system for forward sensitivity. Consider, 
following kinetics:
\begin{align}
    \dot{c}_A &= -kc_A \notag \\
    c_A(0) &= c_{A0}
\end{align}
Say we want to calculate sensitivities wrt $k, c_{A0}$. With reference to \cref{eq:augsens},
for $S_1 = \dfrac{\partial c_A}{\partial k}$ and $S_2 = \dfrac{\partial c_A}{\partial c_{A0}}$,
we have following augmented system:
\begin{align}
    \begin{bmatrix} 
        \dot{c}_A \\
        \dot{S_1} \\ 
        \dot{S_2} 
    \end{bmatrix} = 
    \begin{bmatrix}
        -kc_A \\
        -kS_1 - c_A \\
        -kS_2 \end{bmatrix}
    \qquad
    \begin{bmatrix} 
        c_A(0) \\ 
        S_1(0) \\ 
        S_2(0) 
    \end{bmatrix} =
    \begin{bmatrix} 
        c_{A0} \\ 
        0 \\ 
        1
    \end{bmatrix}
\end{align}
If we were learning the parameter $k$ and initial condition $c_{A0}$, we could
show a gradient descent update as follows:
\begin{align}
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix} \leftarrow
    \begin{bmatrix} k \\ c_{A0} \end{bmatrix}
    - \eta 
    \begin{bmatrix} 
        -2 (\hat{c}_A  - c_A)^T S_1 \\ 
        -2 (\hat{c}_A  - c_A)^T S_2
    \end{bmatrix}
\end{align}

\subsubsection{Backward Sensitivity}

In forward sensitivity since we solve the augmented ODE system, we do not need 
to worry about time-steps taken by the ODE solver. However for backward sensitivity,
it becomes a major problem. Here we show derivation of the adjoint equations for the
continuous version rather than the previous discrete version (where the loss is obtained by summing).
Consider following least-squares objective for \cref{eq:ssode} and the Lagrange functional:

\begin{align}
    V(x) &= \int_0^T (y - \hat{y})^2 dt = \int_0^T g(x, u; \theta)dt \notag \\
    \mathcal{L} &= \int_0^T g dt + \int_0^T \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
\end{align}
Hence,
\begin{align}
    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \dfrac{\partial g}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} dt + 
    \int_0^T \lambda^T \left(\dfrac{\partial f}{\partial x} \dfrac{\partial x}{\partial \theta^T} + \dfrac{\partial f}{\partial \theta} - \dfrac{\partial}{\partial \theta}\dfrac{dx}{dt}\right) dt \notag \\
    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \dfrac{\partial g}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} dt + 
    \int_0^T \lambda^T \left(\dfrac{\partial f}{\partial x} \dfrac{\partial x}{\partial \theta^T} + \dfrac{\partial f}{\partial \theta} - \dfrac{d}{dt}\dfrac{\partial x}{\partial \theta^T}\right) dt \notag \\
    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \lambda^T \dfrac{\partial f}{\partial \theta} dt + \int_0^T \left(\dfrac{\partial g}{\partial x^T}
    + \lambda^T \dfrac{\partial f}{\partial x} - \lambda^T\dfrac{d}{dt}\right)\dfrac{\partial x}{\partial \theta^T} dt
\end{align}
The term $\int_0^T{\lambda^T \dfrac{d}{dt} \dfrac{\partial x}{\partial \theta^T}dt}$ 
can be solved using integration by parts. Thus, we have:

\begin{align}
    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \lambda^T \dfrac{\partial f}{\partial \theta} dt + \int_0^T \left(\dfrac{\partial g}{\partial x^T}
    + \lambda^T \dfrac{\partial f}{\partial x} - \dfrac{d\lambda}{dt}\right)\dfrac{\partial x}{\partial \theta^T} dt - \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0
    + \lambda^T(T)\dfrac{\partial x}{\partial \theta^T}|_T 
    \label{eq:adjoint}
\end{align}

Now, the whole objective of backward sensitivity is to avoid calculating $\dfrac{\partial x}{\partial \theta^T}$.
Also note that the equality constraint in the Langrange functional is always satisfied
since we solve the ODE system. Thus we get to choose $\lambda$ such that the terms
associated with $\dfrac{\partial x}{\partial \theta^T}$ vanish. Also, we can 
easily calculate $\dfrac{\partial x}{\partial \theta^T}|_0$ but we would like to set
$\dfrac{\partial x}{\partial \theta^T}|_T = 0$. Thus, we can construct following adjoint 
equations:

\begin{align}
    \dfrac{d\lambda}{dt} &= \left(\dfrac{\partial g}{\partial x^T} + \lambda^T \dfrac{\partial f}{\partial x}\right)^T \notag \\
    \lambda(T) &= 0
\end{align}
Referring to \cref{eq:adjoint}, we see $\dfrac{\partial V}{\partial \theta} = \dfrac{\partial \mathcal{L}}{\partial \theta} + \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0$.
Thus, we do no need to backprop but just solve the augmented ODE system once forward 
and once backward in each optimal iteration unlike ResNets or RNNs. Also, note that the number 
of adjoint equations do not increase with parameters. However, such a method is not 
an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
numerical ODE solvers like stiffness, accuracy, etc. Sometimes we may end up 
with noisy gradients which may lead to
poor convergence.
